### 2016-11-29 thesis/intro/intro.tex
tmp commit

Though mutation rates can differ between individuals,

``The risk of lung cancer diminishes after smoking cessation, but not during the first 5 years, and the relative risk never turns to that of a nonsmoker'' \citep{Hecht1999}. \\

Tobacco carcinogens form DNA adducts (carcinogen metabolite attached covalently to DNA at A or G), which can be activated or detoxfied in the cell \citep{Hecht1999}. \\

If DNA adducts are not detoxfied, it can lead to DNA miscoding and therefore mutation \citep{Hecht1999}. \\

>>>>

For obesity, since there is a clear evidence of the cause (excessive eating and lack of exercise) and a well defined mechanism of how the disease is developed (excess calories are converted into fat), it is relatively easy to speculate the effect of the environment on the disease (for example, abundance of high calory food).
On the other hand, cancers are developed through many cellular pathways (Hallmarks of Cancers) which are not fully uncovered yet, with either too many or very few causes associated with those pathways (for example, mutations, inflammation and/or growth factors), and therefore it is difficult to assess how, rather than whether, a given environmental factor causes specific cancer.

With that said, there has been a few environmental factors that have a strong evidence of causing cancer, though the complete mechanism of action for some of these must be elucidated in the coming years.
% TODO: do I really need infections and inflammations?
Out of these, infections and inflammation, smoking, \gls{uv} radiation, and diet and obesity will be covered as examples.
\\

% Infections and inflammmation

% Smoking
\noindent
Smoking is a known risk factor for many cancers; the obvious one being lung cancer \citep{Gandini2008,Hecht1999}.
For current smokers, the relative risk is greater than 1.5 in many cancers, including liver, cervix and kidney, and it can rise up to 8.96 in lung cancer \citep{Gandini2008}.
Even if smokers stop smoking, the relative risk does not decrease until the first 5 years have passed and will never return to the level of non-smokers \citep{Hecht1999}.

In tobacco smoke, there are literally hundreds and thousands of chemical compounds, of which about 20 compounds have been shown to have carcinogenic effects \citep{Hecht1999}.
Among these, \glspl{pah} and \gls{nnk} have been shown to cause lung cancer in mice when administered systemically \citep{Hecht1999}.
These carcinogens are processed first by \gls{cyp} enzymes (termed meta\-bolic activation), and the product can either be processed further to be detoxified completely and cause no harm to the cell, or alternatively, the product can covalently bind to the \acrshort{dna}, forming \acrshort{dna} adducts \citep{Hecht1999}.
Unless these adducts are resolved correctly by the \acrshort{dna} repair mechanisms, it can result in a permanent mutation within the genome and lead to oncogenesis \citep{Hecht1999}.

In addition to this, cigarettes contain free radicals and chemicals that cause tremendous oxidative stress and damage to the cells \citep{Hecht1999}.
These radicals cause \acrshort{dna} nicks and damages the \acrshort{dna}, again requring the cell's \acrshort{dna} repair mechanisms to remove the damage so it does not become permanent.
\\

% UV radiation
\noindent
Another important environmental factor to consider is \gls{uv} radiation.
Exposure to \gls{uv} radiation has been significantly associated with the increased risks of skin cancers \citep{Armstrong2001,Gallagher2006}.
The risks of developing any skin cancers can vary between race (Caucasian people are more likely to get skin cancers), and potentially the age when they were exposed to high amounts of \gls{uv} radiation \citep{Armstrong2001,Gallagher2006}.

Again, the mechanism of how \gls{uv} radiation may cause  skin cancers is related to the \acrshort{dna} repair system.
\gls{uv} radiation causes the formation of a dimer in the adjacent pyrimidines, resulting in a covalent bond formation between these nucleotides \citep{Friedberg2003,Hoeijmakers2001}.
These dimers introduce bulky structures in the \acrshort{dna} that prevent the \acrshort{dna} replication and transcription enzymes from functioning properly, and if they are not repaired, it may cause lasting mutations \citep{Friedberg2003,Hoeijmakers2001}.
These damages caused by \gls{uv} radiation can be relieved by the \gls{ner} system or the photoreactivation process, where a photoreactivating enzyme uses light to monomerize the dimeric pyrimidines \citep{Friedberg2003}.
\\

\noindent
As one can imagine, \acrshort{dna} damages caused by smoking and/or \gls{uv} radiation have the potential to damage the genome and initiate cancer if wrong genes are mutated, such as p53 tumour suppressor gene.
Therefore, limited exposure to both factors is crucial for the prevention of any types of cancers, as well as the more specific types of cancers associated with these factors, such as lung and skin cancers.\\

% Diet and obesity
\noindent
